```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.pos = "H",
  out.extra = ""
)
```

```{r book-edition, echo=FALSE, results='asis'}
source("R/book-edition.R")
write_book_edition_script()
```

```{r contact-links, echo=FALSE, results='asis'}
xfun::file_string("contact.html")
```

```{r publication-dates, echo=FALSE, results='asis'}
write_book_dates()
```

## Preface {.unnumbered}

::: {.home-panel}
Option Pricing with R is written for readers who want to connect option pricing, payoffs, option properties, valuation logic, stochastic processes, and reproducible R code. The book focuses on the mechanics that make option pricing work: payoff diagrams, put-call parity, binomial valuation, risk-neutral reasoning, Black-Scholes intuition, and the Wiener-process foundations behind continuous-time models.

The emphasis is applied and computational. Each chapter keeps the financial object visible while the code is built, so that option values, bounds, probabilities, paths, and valuation relationships can be inspected numerically and graphically. Future revisions will extend this foundation toward Greeks and richer option-pricing applications.
:::

## What's new in this edition {.unnumbered}

- **Sharper book identity:** The book now focuses on option pricing, stochastic processes, and valuation mechanics.
- **Cleaner reading flow:** Payoff functions, option properties, and Wiener processes open as separate chapters with local section menus.
- **Publication workflow:** The site is prepared for GitHub Pages through GitHub Actions, with `_book` treated as generated output.
